---
title: "Melbourne Rain Prediction"
author: "Rahul K Gupta (s3635232) & Terrie Christensen (s3664899)"
date: "7 April 2018"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: '3'
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
linkcolor: blue
subtitle: MATH 2319 Machine Learning Applied Project Phase I
documentclass: article
---

\newpage

\tableofcontents

\newpage

# Introduction \label{sec1}

The objective of this project was to build classifiers to predict whether an individual earns more than USD 50,000 or less in a year from the 1994 US Census Data. The data sets were sourced from the [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Census+Income). This project has two phases. Phase I focuses on data preprocessing and exploration, as covered in this report. We shall present model building in Phase II. The rest of this report is organised as follow. Section 2 describes the data sets and their attributes. Section 3 covers data pre-processing. In Section 4, we explore each attribute and their inter-relationships. The last section ends with a summary.

# Data Set

The [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Census+Income) provides five data sets, but only `adult.data`, `adult.test`, and `adult.names` were useful in this project. `adult.data` and `adult.test` are the training and test data sets respectively. `adult.names` contains the details of attributes or variables. The training data set has 32,561 training observations. Meanwhile, the test data set has 16,281 test observations. Both data sets consist of 14 descriptives features and one target feature. In this project, we combined both training and test data into one. In Phase II, we would build the classifiers from the combined the data set and evaluate their performance using cross-validation.

## Target Feature

The response feature of rain is given as:

\begin{center}
  \begin{align}
    \text{Tomorrow's Rain} & = 
      \begin{cases} 
         Yes & \text{if Rain will occur tomorrow} \\
         No & \text{otherwise}
      \end{cases}
  \end{align}
\end{center}

The target feature has two classes and hence it is a binary classification problem. To reiterate, The goal is to predict **whether it will rain in Melbourne tomorrow**.

## Descriptive Features

The variable description is produced here from `adult.names` file:

* age: continuous.
* workclass: Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, State-gov, Without-pay,
Never-worked.
* fnlwgt: continuous.
* education: Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, Assoc-voc, 9th, 7th-
8th, 12th, Masters, 1st-4th, 10th, Doctorate, 5th-6th, Preschool.
* education-num: continuous.
* marital-status: Married-civ-spouse, Divorced, Never-married, Separated, Widowed, Married-
spouse-absent, Married-AF-spouse.
* occupation: Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, Prof-specialty,
Handlers-cleaners, Machine-op-*inspct, Adm-clerical, Farming-fishing, Transport-moving, Priv-
house-serv, Protective-serv, Armed-Forces.
* relationship: Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.
* race: White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.
* sex: Female, Male.
* capital-gain: continuous.
* capital-loss: continuous.
* hours-per-week: continuous.
* native-country: United-States, Cambodia, England, Puerto-Rico, Canada, Germany, Outlying-
US(Guam-USVI-etc), India, Japan, Greece, South, China, Cuba, Iran, Honduras, Philippines, Italy, Poland, Jamaica, Vietnam, Mexico, Portugal, Ireland, France, Dominican-Republic, Laos, Ecuador, Taiwan, Haiti, Columbia, Hungary, Guatemala, Nicaragua, Scotland, Thailand, Yugoslavia, El Salvador, Trinadad&Tobago, Peru, Hong, Holand-Netherlands.

Most of the descriptive features are self-explanatory, except \textcolor{red}{fnlwgt} which stands for "Final Weight" defined by the US Census. The weight is an "estimate of the number of units in the target population that the responding unit represents". This feature aims to allocate similar weights to people with similar demographic characteristics. For more details, see [US Census](https://www.census.gov/programs- surveys/sipp/methodology/weighting.html).

# Data Pre-processing


## Preliminaries (Optional)

In this project, we used the following `R` packages.

```{r global_options, include=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r}
library(knitr)
library(mlr)
library(tidyverse)
library(GGally)
library(cowplot)
require(corrplot)
library(reshape2)
theme_set(theme_minimal())
theme_weather <- theme_minimal() +
  theme(plot.subtitle = element_text(color = '#333333', face = "italic"),
        plot.caption = element_text(color = '#666666', face = "italic", size = 9))
```

Text

```{r}
weather <- read.csv('weatherAUS 2.csv')
weather$Date = as.Date(weather$Date,'%Y-%m-%d')

# Filter data for Melbourne
Melbourne = weather[weather$Location %in% c('Melbourne','MelbourneAirport'),]

# Add Month and Year-Month
Melbourne$Month = strftime(Melbourne$Date,"%b")
Melbourne$MonthYear = strftime(Melbourne$Date,"%b-%Y")
```


## Data Cleaning and Transformation

Evaluating missing values via visualizations

```{r}
# Missing Year Data
missing_year = subset(Melbourne ,is.na(Melbourne$RainToday) | is.na(Melbourne$RainTomorrow))
missing_year$Year = strftime(missing_year$Date,"%Y")
ggplot(missing_year,aes(Year)) + 
  geom_bar(fill='#969696') +
  ggtitle('Melbourne Weather Missing Data by Year') +
  labs(subtitle = '2015 has most number of Missing Data',
       caption="Source - Commonwealth of Australia , Bureau of Meteorology") +
  theme_weather
```

```{r}
# Missing Month Data 
missing_month = subset(Melbourne ,is.na(RainToday) | is.na(RainTomorrow)) 

missing_month$Month = factor(missing_month$Month,
                             levels = c('Jan','Feb','Mar','Apr','May','Jun',
                                        'Jul','Aug','Sep','Oct','Nov','Dec'), 
                             ordered = TRUE)

ggplot(missing_month,aes(Month)) + 
  geom_bar(fill='#969696') +
  ggtitle('Melbourne Weather Missing Data by Month') +
  labs(subtitle = 'Apr has highest missing values, Dec has least',
       caption="Source - Commonwealth of Australia , Bureau of Meteorology") +
  theme_weather
# Since the month data is distributed well, we can remove them from data set
```

```{r}
# Missing data percentage
missing_percentage = count(missing_month)/count(Melbourne) * 100
missing_percentage

# Remove data where Today and Tomorrow rain information is missing
# 14.4% data is missing
```

```{r}
# Remove Missing Data
Melbourne = Melbourne %>% 
  drop_na(RainToday) %>%
  drop_na(RainTomorrow)
```


With `str` and `summarizeColumns` (see Table 1), we noticed the following anomalies:

* All character columns contained excessive white space.
* The target feature, `income` had a cardinality of 4, which was supposed to be 2 since `income` must be binary.
* The `education_num` ranged from 1 to 16 which coincided with the cardinality of `education`. They might represent the same information.
* The max value of `capital_gain` was 99999, potentially a value to represent missing value.
* The max value of `hours_per_week` was 99. It could be a valid or missing value
* On surface, each feature had no missing value, especially the character features.

```{r}
str(Melbourne)
```

```{r}
summarizeColumns(Melbourne) %>% knitr::kable(caption =  'Feature Summary ')
```

```{r}
# Change cardinal direction to degrees
direction = read.csv('direction.csv')
Melbourne$WindGustDirDegree = factor(Melbourne$WindGustDir, 
                               levels = direction$CardinalDirection, 
                               labels = direction$DegreeDirectionMean)

Melbourne$WindDir9amDegree = factor(Melbourne$WindDir9am, 
                               levels = direction$CardinalDirection, 
                               labels = direction$DegreeDirectionMean)

Melbourne$WindDir3pmDegree = factor(Melbourne$WindDir3pm, 
                               levels = direction$CardinalDirection, 
                               labels = direction$DegreeDirectionMean)
```


## Univariate Visualisation



```{r}
# --------------
# Visualizations
# --------------

# ------------------
# Monthly Average Rainfall from 2009-2016
total_yearmonth_rainfall = Melbourne %>%
  select(MonthYear,Rainfall) %>%
  group_by(MonthYear) %>%
  summarise(monthTotal = sum(Rainfall, na.rm = TRUE))

total_yearmonth_rainfall$month = substr(total_yearmonth_rainfall$MonthYear, 1, 3)

monthly_rainfall =  total_yearmonth_rainfall %>%
  group_by(month) %>%
  summarise(avgRainfall = mean(monthTotal))

monthly_rainfall$month = factor(monthly_rainfall$month,
                                levels = c('Jan','Feb','Mar','Apr',
                                           'May','Jun','Jul','Aug',
                                           'Sep','Oct','Nov','Dec'),
                                ordered = TRUE)

ggplot(monthly_rainfall,aes(month,avgRainfall)) +
  geom_bar(stat = 'identity',fill="#006699") +
  ggtitle('Monthly Average Rainfall, 2009-2016') +
  xlab('Months') +
  ylab('Rainfall(MM)') +
  labs(subtitle = 'Rainfall is increasing from Jan to Dec with little variation') +
  theme(plot.subtitle = element_text(color = '#333333',face = "italic"))

# Except Nov, the variation isn't too much between Consecutive years. The possibility of getting rain 
# in Feb is almost same as Mar or Apr, hence we connot conclude the rain by month. 
# Melbourne doesn't have a fix rainy season hence average rainfall is distributed
```

### Correlation Matrix

```{r}
# ------------------
# Correlation Matrix
correlation_matrix = cor(Melbourne[,c(3:7,9,12:21)],use='na.or.complete')
corrplot(correlation_matrix, order = 'AOE', type = "upper")
# Very few factors has high correlation

```


### continuous variables

```{r}

# visualize continuous variables
continouous_cols = c('MinTemp','MaxTemp','Rainfall','Evaporation','Sunshine',
                     'WindGustSpeed','WindSpeed9am','WindSpeed3pm','Humidity9am',
                     'Humidity3pm','Pressure9am','Pressure3pm','Temp9am','Temp3pm')
continouous_variables = melt(Melbourne[,continouous_cols])

ggplot(continouous_variables) + 
  stat_bin(aes(value), color='#EEEEEE', size=.1, fill="#006699") + 
  facet_wrap(~variable,scales="free") +
  ggtitle('Melbourne Weather - Feature Distribution') +
  xlab('') +
  ylab('') +
  labs(subtitle = 'Temperature and Pressure is normally distributed',
       caption="Source - Commonwealth of Australia , Bureau of Meteorology") +
  theme_weather
```


### Categorical variables

```{r}
# visualize categorical variables
head(Melbourne)

wind_direction = Melbourne 
wind_direction$WindGustDir = factor(wind_direction$WindGustDir, 
                       levels = rev(direction$CardinalDirection), 
                       ordered = TRUE)

wind_direction %>%
  filter(!is.na(WindGustDir)) %>%
  ggplot(aes(WindGustDir)) + 
  geom_bar(fill="#006699") + 
  coord_flip() +
  ggtitle('Melbourne Direction of the Strongest Wind Gust in 24 Hours') +
  ylab('Count') +
  xlab('16 compass points') +
  labs(subtitle = 'Direction is ordered from 0 degrees North in clockwise order',
       caption="Source - Commonwealth of Australia , Bureau of Meteorology") +
  theme_weather



```



```{r}
  

  melt(Melbourne[,c('Cloud9am','Cloud3pm')]) %>%
  ggplot(aes(value)) + 
  geom_bar() + 
  facet_wrap(~variable,scales="free") +
  coord_flip()




# ---------------------------------------------------------
# Relationship Between Rain and Minimum Maximum Temperature
# Convert wide to long
Temperature = Melbourne %>% 
  select(MinTemp,MaxTemp,RainToday) %>%
  drop_na(RainToday) %>%
  melt(id.vars = 'RainToday')

ggplot(Temperature,aes(variable,value)) + 
  geom_boxplot(fill=c('#9ecae1','#3182bd','#9ecae1','#3182bd')) + 
  facet_grid(~RainToday) +
  theme_gray() +
  ggtitle('Relationship Between Rain and Minimum Maximum Temperature') +
  xlab('Months') +
  ylab('Temperature(Celsius)') +
  labs(subtitle = 'Difference between High and low temperature is less on Rainy day') +
  theme(plot.subtitle = element_text(color = '#333333',face = "italic"))


```


### Monthly Average Rainfall from 2009-2016

```{r}
# ------------------
# Monthly Average Rainfall from 2009-2016

# Not working

# Except Nov, the variation isn't too much between Consecutive years. The possibility of getting rain 
# in Feb is almost same as Mar or Apr, hence we connot conclude the rain by month. 
# Melbourne doesn't have a fix rainy season hence average rainfall is distributed

```


```{r}



#################
# Delta change
#################

delta_cloud = Melbourne
delta_cloud$deltaCloud = delta_cloud$Cloud9am - delta_cloud$Cloud3pm
ggplot(delta_cloud, aes(deltaCloud, fill=RainToday)) + 
  geom_bar(position = 'dodge') +
  ggtitle('Melbourne Weather - Cloud Delta Change in Oktas') +
  xlab('Oktas') +
  ylab('') +
  labs(subtitle = '0 is completely clear sky and 8 is completely overcast',
       caption="Source - Commonwealth of Australia , Bureau of Meteorology") +
  theme_weather

```


### Comparision

```{r}
# ----------------------------------------------
# Temperature frop from 9AM to 3 PM for the same day 
temperature_drop= Melbourne[Melbourne$Temp9am < Melbourne$Temp3pm, ]
prop.table(table(temperature_drop$RainToday)) * 100
# Even if temperature drops by 3 PM, chances of having rain is just 22% 
temperature_drop= Melbourne[Melbourne$Temp9am > Melbourne$Temp3pm, ]
prop.table(table(temperature_drop$RainToday)) * 100
# ----------------------------------------------


# ----------------------------------------------
# Relationship between today and tomorrow's rain
prop.table(table(Melbourne$RainToday, Melbourne$RainTomorrow,dnn=c('Rain Today','Rain Tomorrow'))) * 100
# 9% times if it rains today then it rained tomorrow 
# 14% times it rains today and didn't rain tomorrow
# If it didn't rain today then chances are it won't rain tomorrow

# TODO we can also do chi square test of independence
# ----------------------------------------------
```





# Summary

