---
title: "Melbourne Rain Prediction"
author: "Rahul K Gupta (s3635232) & Terrie Christensen (s3664899)"
date: "7 April 2018"
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  html_document:
    df_print: paged
    toc: no
    toc_depth: '3'
linkcolor: blue
subtitle: MATH 2319 Machine Learning Applied Project Phase I
documentclass: article
---

\newpage

\tableofcontents

\newpage

# Introduction \label{sec1}

The objective of this project was to build classifiers to predict whether an individual earns more than USD 50,000 or less in a year from the 1994 US Census Data. The data sets were sourced from the [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Census+Income). This project has two phases. Phase I focuses on data preprocessing and exploration, as covered in this report. We shall present model building in Phase II. The rest of this report is organised as follow. Section 2 describes the data sets and their attributes. Section 3 covers data pre-processing. In Section 4, we explore each attribute and their inter-relationships. The last section ends with a summary.

# Data Set

The [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Census+Income) provides five data sets, but only `adult.data`, `adult.test`, and `adult.names` were useful in this project. `adult.data` and `adult.test` are the training and test data sets respectively. `adult.names` contains the details of attributes or variables. The training data set has 32,561 training observations. Meanwhile, the test data set has 16,281 test observations. Both data sets consist of 14 descriptives features and one target feature. In this project, we combined both training and test data into one. In Phase II, we would build the classifiers from the combined the data set and evaluate their performance using cross-validation.

## Target Feature

The response feature of rain is given as:

\begin{center}
  \begin{align}
    \text{Tomorrow's Rain} & = 
      \begin{cases} 
         Yes & \text{if Rain will occur tomorrow} \\
         No & \text{otherwise}
      \end{cases}
  \end{align}
\end{center}

The target feature has two classes and hence it is a binary classification problem. To reiterate, The goal is to predict **whether it will rain in Melbourne tomorrow**.

## Descriptive Features

The variable description is produced here from `adult.names` file:

* age: continuous.
* workclass: Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, State-gov, Without-pay,
Never-worked.
* fnlwgt: continuous.
* education: Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, Assoc-voc, 9th, 7th-
8th, 12th, Masters, 1st-4th, 10th, Doctorate, 5th-6th, Preschool.
* education-num: continuous.
* marital-status: Married-civ-spouse, Divorced, Never-married, Separated, Widowed, Married-
spouse-absent, Married-AF-spouse.
* occupation: Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, Prof-specialty,
Handlers-cleaners, Machine-op-*inspct, Adm-clerical, Farming-fishing, Transport-moving, Priv-
house-serv, Protective-serv, Armed-Forces.
* relationship: Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.
* race: White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.
* sex: Female, Male.
* capital-gain: continuous.
* capital-loss: continuous.
* hours-per-week: continuous.
* native-country: United-States, Cambodia, England, Puerto-Rico, Canada, Germany, Outlying-
US(Guam-USVI-etc), India, Japan, Greece, South, China, Cuba, Iran, Honduras, Philippines, Italy, Poland, Jamaica, Vietnam, Mexico, Portugal, Ireland, France, Dominican-Republic, Laos, Ecuador, Taiwan, Haiti, Columbia, Hungary, Guatemala, Nicaragua, Scotland, Thailand, Yugoslavia, El Salvador, Trinadad&Tobago, Peru, Hong, Holand-Netherlands.

Most of the descriptive features are self-explanatory, except \textcolor{red}{fnlwgt} which stands for "Final Weight" defined by the US Census. The weight is an "estimate of the number of units in the target population that the responding unit represents". This feature aims to allocate similar weights to people with similar demographic characteristics. For more details, see [US Census](https://www.census.gov/programs- surveys/sipp/methodology/weighting.html).

# Data Pre-processing


## Preliminaries (Optional)

In this project, we used the following `R` packages.

```{r global_options, include=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r}
library(mlr)
library(tidyverse)
require(corrplot)
library(reshape2)
theme_set(theme_minimal())
theme_weather <- theme_minimal() +
  theme(plot.subtitle = element_text(color = '#333333',face = "italic"))
```

Text

```{r}
weather <- read.csv('weatherAUS 2.csv')
weather$Date = as.Date(weather$Date,'%Y-%m-%d')

# Filter data for Melbourne
Melbourne = weather[weather$Location %in% c('Melbourne','MelbourneAirport'),]

# Add Month and Year-Month
Melbourne$Month = strftime(Melbourne$Date,"%b")
Melbourne$MonthYear = strftime(Melbourne$Date,"%b-%Y")

# Remove data where Tomorrow rain information is missing
# Melbourne = Melbourne %>% drop_na(RainTomorrow)
```


## Data Cleaning and Transformation

With `str` and `summarizeColumns` (see Table 1), we noticed the following anomalies:

* All character columns contained excessive white space.
* The target feature, `income` had a cardinality of 4, which was supposed to be 2 since `income` must be binary.
* The `education_num` ranged from 1 to 16 which coincided with the cardinality of `education`. They might represent the same information.
* The max value of `capital_gain` was 99999, potentially a value to represent missing value.
* The max value of `hours_per_week` was 99. It could be a valid or missing value
* On surface, each feature had no missing value, especially the character features.

```{r}
str(Melbourne)
```

```{r}
summarizeColumns(Melbourne) %>% knitr::kable(caption =  'Feature Summary ')
```

## Univariate Visualisation

### Sub Headings




# Summary

